---
title: "p8105_hw3_sjz2120"
author: "Sophie Zhang (sjz2120)"
date: "2022-10-09"
output: github_document
---
```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%")

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


# Problem 1
This problem uses the Instacart data. DO NOT include this dataset in your local data directory; instead, load the data from the `p8105.datasets` using:

```{r, "Loading in Instacart dataset"}
library(p8105.datasets)
data("instacart")

instacart = 
  instacart %>% 
  as_tibble(instacart)
```

### Describing the Instacart dataset

This dataset contains a total of `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row representing a single product from an instacart order.

Key variables include identifiers for each user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, as well as number of days since prior order. There are also several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether or not the item has been ordered by the specific  user in the past.

In total, there are `r instacart %>% select(product_id) %>% distinct %>% count` products found in `r instacart %>% select(user_id, order_id) %>% distinct %>% count` orders from `r instacart %>% select(user_id) %>% distinct %>% count` distinct users.


### Most popular aisles

* How many aisles are there, and which aisles are the most items ordered from?

**There are 134 aisles in total, with most items ordered from the fresh vegetables aisle followed closely by the fresh fruits aisle.**

```{r}
instacart %>% 
  count(aisle) %>% 
  arrange(desc(n))
```

### Plotting the number of items ordered in each aisle

This plot shows the number of items ordered in each aisle, limited to aisles with more than 10000 items ordered. Aisles are arranged in ascending order of the number of items ordered in that aisle:
```{r}
instacart %>% 
  count(aisle) %>% 
  filter(n > 10000) %>% 
  mutate(aisle = fct_reorder(aisle, n)) %>% 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


### Table of most popular items from aisles: “baking ingredients”, “dog food care”, and “packaged vegetables fruits”

This table shows the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. The number of times each item is ordered is also included in this table, and the rows are ordered by descending number of times each item is ordered.
```{r}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle) %>% 
  count(product_name) %>% 
  mutate(rank = min_rank(desc(n))) %>% 
  filter(rank < 4) %>% 
  arrange(desc(n)) %>%
  knitr::kable()
```

### Table of average hours of the day when Pink Lady Apples and Coffee Ice Cream are ordered, on each day of the week

* This final table shows the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week, formatted as a 2x7 table.
```{r}
instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day)) %>%
  spread(key = order_dow, value = mean_hour) %>%
  knitr::kable(digits = 2)
```



# Problem 2
Accelerometers have become an appealing alternative to self-report techniques for studying physical activity in observational studies and clinical trials, largely because of their relative objectivity. During observation periods, the devices measure “activity counts” in a short period; one-minute intervals are common. Because accelerometers can be worn comfortably and unobtrusively, they produce around-the-clock observations.

This problem uses five weeks of accelerometer data collected on a 63 year-old male with BMI 25, who was admitted to the Advanced Cardiac Care Center of Columbia University Medical Center and diagnosed with congestive heart failure (CHF). In this spreadsheet, variables activity. are the activity counts for each minute of a 24-hour day starting at midnight.


### Loading, tidying, and wrangling the accelerometer dataset `accel_data`
Loading, tidying, and otherwise wrangleingthe data. The final dataset `accel_data` includes all originally observed variables and values; has useful variable names; includes a weekday vs weekend variable; and encodes data with reasonable variable classes.

```{r, "Load, tidy, and wrangle Problem 2 Accelerometer data"}
accel_data =
  read_csv(file = "./data/accel_data.csv") %>% #Read in csv file
  janitor::clean_names() %>% #Basic clean of dataset
  rename_with(.cols = starts_with("activity_"),
              .fn = ~ str_replace(string = .x, "activity_", "minute_")) %>% #Renamed all activity_* variables to minute_*
  mutate(type_of_day = recode(day, "Monday" = "Weekday", "Tuesday" = "Weekday", "Wednesday" = "Weekday", "Thursday" = "Weekday", "Friday" = "Weekday", "Saturday" = "Weekend", "Sunday" = "Weekend" )) %>% #Create new variable indicating weekday vs weekend
  select(week, day_id, day, type_of_day, everything()) #Reorganize columns so minute_* are at the end
```

### Describing the cleaned & wrangled `accel_data` dataset

The cleaned and tidied dataset `accel_data` contains a total of `r nrow(accel_data)` rows and `r ncol(accel_data)` columns, with each row representing a single day of accelerometer data (for each minute) for the subject. There are a total of 5 weeks of data, with 7 days of data per week for a total of `r nrow(accel_data)` total days (days when accelerometer observations occured), with accelerometer data for each minute of the day.

Key variables include `week` (the week during which the row data was collected), `day_id` (an identifier for each of the `r nrow(accel_data)` days that data was collected for), and `day` (a label name for the day of the week); as well as a newly created variable `type_of_day` (which indicates whether the observation happened on a weekday vs weekend).

Other key variables include `minute_*` (renamed from `activity_*`) and ranging from `minute_1` to `minute_1440`. The variables `minute_*` are the activity counts for each minute of a 24-hour day starting at midnight, hence why there are a total of 1440 `minute_*` variables (because there are 1440 minutes in a day).

All of these variables are doubles, with the exception of the `day` and `type_of_day` variables which are character variable types.


### Table with aggregated total daily activity per each day (`accel_data`)

Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

```{r, "Table with aggregated daily total accelerometer activity"}
accel_data %>%
  mutate(daily_activity = rowSums(accel_data[ , c(5:1444)], na.rm = TRUE)) %>%
  select(week, day_id, day, type_of_day, daily_activity) %>%
  knitr::kable()
```

**There don't seem to be any noticeable trends in the aggregated amount of daily activity (`daily_activity`). However, there do seem to be a few days (without any noticeable patterns as to which days, except for the two Saturdays in weeks 4 and 5) where daily activity is significantly lower compared to other days.**


### Single-panel plot with 24-hour activity time courses for each day (`accel_data`)

Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.

```{r}
accel_pivot_data = 
  pivot_longer(
    accel_data, 
    minute_1:minute_1440,
    names_to = "minute", 
    names_prefix = "minute_",
    values_to = "activity_level") %>% #Create new dataframe from accel_data that has activity_level as it's own variable
  mutate(hour = as.numeric(minute)/60, day_id = as.character(day_id))



ggplot(accel_pivot_data, aes(x = hour, y = activity_level, color = day)) +
  geom_smooth(se = FALSE)



```



# Problem 3
This problem uses the NY NOAA data. DO NOT include this dataset in your local data directory; instead, load the data from the `p8105.datasets` package using:

```{r, "Loading in Problem 3 dataset"}
library(p8105.datasets)
data("ny_noaa")

noaa = 
  ny_noaa %>% 
  as_tibble(ny_noaa)

noaa
```


### Describing the NY NOAA (`noaa`) dataset

**Background**: The National Oceanic and Atmospheric Association (NOAA) provides public access to some weather data, including the GHCN (Global Historical Climatology Network)-Daily database of summary statistics from weather stations around the world. GHCN-Daily contains records from over 100,000 stations in 180 countries and territories. NCEI provides numerous daily variables, including maximum and minimum temperature, total daily precipitation, snowfall, and snow depth; however, about one half of the stations report precipitation only. Both the record length and period of record vary by station and cover intervals ranging from less than a year to more than 175 years.

**Describing the Dataset**: The newly loaded `noaa` dataset contains a total of `r nrow(noaa)` rows and `r ncol(noaa)` columns, with each row representing a single day of observations/weather data for a specific weather station. There are a total of `r noaa %>% count(id) %>% nrow()` unique weather stations from where weather data was collected, and the dates of the observations range from `r noaa %>% pull(date) %>% min()` to `r noaa %>% pull(date) %>% min()`.

Key identifier variables include `id` (the weather station ID associated with the specific row of data and `date` (date of observation). Other key variables included the various weather measures such as `prcp` (Precipitation (tenths of mm)), `snow` (Snowfall (mm)), `snwd` (Snow depth (mm)), `tmax` (Maximum temperature (tenths of degrees C)), and `tmin` (Minimum temperature (tenths of degrees C)).

Some of the rows/observations have missing values. In fact, let's see how many missing values there are for each weather data variable:

* There are `r sum(is.na(noaa$prcp))` observations with missing values for the variable `prcp` (Precipitation (tenths of mm))

* There are `r sum(is.na(noaa$snow))` observations with missing values for the variable `snow` (Snowfall (mm))

* There are `r sum(is.na(noaa$snwd))` observations with missing values for the variable `snwd` (Snow depth (mm))

* There are `r sum(is.na(noaa$tmax))` observations with missing values for the variable `tmax` (Maximum temperature (tenths of degrees C)) -- that's nearly half of all observations that have missing values for `tmax`!

* There are `r sum(is.na(noaa$tmin))` observations with missing values for the variable `tmin` (Minimum temperature (tenths of degrees C)) -- that's nearly half of all observations that have missing values for `tmin`!

* There are no observations with missing values for the variables `id` and `date` (thankfully!)


### Data cleaning the NY NOAA (`noaa`) dataset

Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?


### Two-panel plot of average max temperature in January and July, in each station across years (`noaa` dataset)

Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?


### Two-panel plot of tmax and tmin (`noaa` dataset)

Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option)


### Two-panel plot of snowfall distribution, separately by year (`noaa` dataset)

Make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

