---
title: "p8105_hw3_sjz2120"
author: "Sophie Zhang (sjz2120)"
date: "2022-10-09"
output: github_document
---
```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%")

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


# Problem 1
This problem uses the Instacart data. DO NOT include this dataset in your local data directory; instead, load the data from the `p8105.datasets` using:

```{r Loading_in_Instacart_data}
library(p8105.datasets)
data("instacart")

instacart = 
  instacart %>% 
  as_tibble(instacart)
```

### Describing the Instacart dataset

This dataset contains a total of `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row representing a single product from an instacart order.

Key variables include identifiers for each user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, as well as number of days since prior order. There are also several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether or not the item has been ordered by the specific  user in the past.

In total, there are `r instacart %>% select(product_id) %>% distinct %>% count` products found in `r instacart %>% select(user_id, order_id) %>% distinct %>% count` orders from `r instacart %>% select(user_id) %>% distinct %>% count` distinct users.


### Most popular aisles

* How many aisles are there, and which aisles are the most items ordered from?

**There are 134 aisles in total, with most items ordered from the fresh vegetables aisle followed closely by the fresh fruits aisle.**

```{r Counting_aisles}
instacart %>% 
  count(aisle) %>% 
  arrange(desc(n))
```

### Plotting the number of items ordered in each aisle

This plot shows the number of items ordered in each aisle, limited to aisles with more than 10000 items ordered. Aisles are arranged in ascending order of the number of items ordered in that aisle:
```{r Plotting_aisle_items}
instacart %>% 
  count(aisle) %>% 
  filter(n > 10000) %>% 
  mutate(aisle = fct_reorder(aisle, n)) %>% 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


### Table of most popular items from aisles: “baking ingredients”, “dog food care”, and “packaged vegetables fruits”

This table shows the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. The number of times each item is ordered is also included in this table, and the rows are ordered by descending number of times each item is ordered.
```{r Table_of_items_ordered}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle) %>% 
  count(product_name) %>% 
  mutate(rank = min_rank(desc(n))) %>% 
  filter(rank < 4) %>% 
  arrange(desc(n)) %>%
  knitr::kable()
```

### Table of average hours of the day when Pink Lady Apples and Coffee Ice Cream are ordered, on each day of the week

* This final table shows the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week, formatted as a 2x7 table.
```{r Table_mean_hours}
instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day)) %>%
  spread(key = order_dow, value = mean_hour) %>%
  knitr::kable(digits = 2)
```



# Problem 2
Accelerometers have become an appealing alternative to self-report techniques for studying physical activity in observational studies and clinical trials, largely because of their relative objectivity. During observation periods, the devices measure “activity counts” in a short period; one-minute intervals are common. Because accelerometers can be worn comfortably and unobtrusively, they produce around-the-clock observations.

This problem uses five weeks of accelerometer data collected on a 63 year-old male with BMI 25, who was admitted to the Advanced Cardiac Care Center of Columbia University Medical Center and diagnosed with congestive heart failure (CHF). In this spreadsheet, variables activity. are the activity counts for each minute of a 24-hour day starting at midnight.


### Loading, tidying, and wrangling the accelerometer dataset `accel_data`
Loading, tidying, and otherwise wrangleingthe data. The final dataset `accel_data` includes all originally observed variables and values; has useful variable names; includes a weekday vs weekend variable; and encodes data with reasonable variable classes.

```{r, Load_Problem2_accelerometer_data"}
accel_data =
  read_csv(file = "./data/accel_data.csv") %>% # Read in csv file
  janitor::clean_names() %>% #Basic clean of dataset
  pivot_longer(activity_1:activity_1440, # Pivot longer to condense the activity variables into a single new variable, but still keep all of the observations and variable indicators
               names_to = "minutes_after_midnight",
               names_prefix = "activity_",
               values_to = "activity") %>%
  mutate(type_of_day = if_else((day == "Saturday" | day == "Sunday"), "Weekend", "Weekday"),
         day = factor(day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"), 
                      ordered = TRUE),
         minutes_after_midnight = as.numeric(minutes_after_midnight)) %>%
  select(week, day_id, day, type_of_day, everything()) # Reorder the variables so day is next to type_of_day

accel_data
```

### Describing the cleaned & wrangled `accel_data` dataset

The cleaned and tidied dataset `accel_data` contains a total of `r nrow(accel_data)` rows/observations and `r ncol(accel_data)` columns/variables, with each row representing a single minute of accelerometer data for the subject for a total of `r nrow(accel_data)` minutes for which they had accelerometer observation data. Each row/observation has a week identifier and day identifier for when the observation was taken; as well as a `minutes_after_midnight` identifier (the number of minutes after midnight when the observation was made).

Key variables include `week` (the week during which the observation was collected), `day_id` (an identifier for each of the `r nrow(accel_data)` days that data was collected for), and `day` (a label name for the day of the week during which the observation was collected); as well as a newly created variable `type_of_day` (which indicates whether the observation happened on a weekday vs weekend).

Other key variables include `minutes_after_midnight` (derived from the previous 1440 `activity_*` variables). The variable `minutes_after_midnight` is the number of minutes from midnight when the observation was collected. Another key variable `activity` describes the amount of activity recorded by the accelerometer for that minute of observation.

All of these variables are doubles, with the exception of the `day` and `type_of_day` variables which are character variable types.


### Table with aggregated total daily activity per each day (`accel_data`)

Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

```{r, Table_of_aggregated_daily_accel_activity"}
accel_data %>%
  group_by(week, day) %>%
  summarise(daily_activity = sum(activity)) %>%
  pivot_wider(names_from = day,
              values_from = daily_activity) %>%
  knitr::kable()
```

**In general, it seems like for each week total daily activity increases from Monday to Friday, and decreases/drops slightly on the weekend (Saturday and Sunday). There also seem to be a few days (without any noticeable patterns as to which days, except for the two Saturdays in weeks 4 and 5) where daily activity is significantly lower compared to other days.**


### Single-panel plot with 24-hour activity time courses for each day (`accel_data`)

Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.

We use `minutes_after_midnight`/60 as the x variable, since we want to show time on a scale of hours and not minutes; our y-variable is activity. We can then make a 24-hour activity time course plot using geom_line:

```{r}
accel_data %>%
  ggplot(aes(x = minutes_after_midnight/60, y = activity)) + # Divide minutes_after_midnight by 60 to get time in hour units
  geom_line(aes(color = day), alpha = 0.5, se = FALSE) +
  labs(title = "24-hour time course plot of activity, by day of the week",
       x = "Hour (after midnight)",
       y = "Activity Level (accelerometer measurements)") +
  theme_classic()
```

From this graph, we can see that there is generally low/much less activity from hours 0 to 5 (most likely because the subject is sleeping). Activity levels are relatively higher from hours 5 to 19 (probably when the subject is awake), and activity levels generally seem to peak from hours 20 to 22 for most days of the week, with the exception of Sunday (Sunday peak activity hours seem to be around 10 to 11 hours after midnight). Activity levels for all days begin to drop after hour 22, and continue to be relatively low until the next day.

# Problem 3
This problem uses the NY NOAA data. DO NOT include this dataset in your local data directory; instead, load the data from the `p8105.datasets` package using:

```{r Loading_in_Problem3_dataset_noaa"}
library(p8105.datasets)
data("ny_noaa")

noaa = 
  ny_noaa
```


### Describing the NY NOAA (`noaa`) dataset

**Background**: The National Oceanic and Atmospheric Association (NOAA) provides public access to some weather data, including the GHCN (Global Historical Climatology Network)-Daily database of summary statistics from weather stations around the world. GHCN-Daily contains records from over 100,000 stations in 180 countries and territories. NCEI provides numerous daily variables, including maximum and minimum temperature, total daily precipitation, snowfall, and snow depth; however, about one half of the stations report precipitation only. Both the record length and period of record vary by station and cover intervals ranging from less than a year to more than 175 years.

**Describing the Dataset**: The newly loaded `noaa` dataset contains a total of `r nrow(noaa)` rows and `r ncol(noaa)` columns, with each row representing a single day of observations/weather data for a specific weather station. There are a total of `r noaa %>% count(id) %>% nrow()` unique weather stations from where weather data was collected, and the dates of the observations range from `r noaa %>% pull(date) %>% min()` to `r noaa %>% pull(date) %>% min()`.

Key identifier variables include `id` (the weather station ID associated with the specific row of data and `date` (date of observation). Other key variables included the various weather measures such as `prcp` (Precipitation (tenths of mm)), `snow` (Snowfall (mm)), `snwd` (Snow depth (mm)), `tmax` (Maximum temperature (tenths of degrees C)), and `tmin` (Minimum temperature (tenths of degrees C)).

Some of the rows/observations have missing values. In fact, let's see how many missing values there are for each weather data variable:

* There are `r sum(is.na(noaa$prcp))` observations (proportion: `r sum(is.na(noaa$prcp))/nrow(noaa)`) with missing values for the variable `prcp` (Precipitation (tenths of mm))

* There are `r sum(is.na(noaa$snow))` observations (proportion: `r sum(is.na(noaa$snow))/nrow(noaa)`) with missing values for the variable `snow` (Snowfall (mm))

* There are `r sum(is.na(noaa$snwd))` observations (proportion: `r sum(is.na(noaa$snwd))/nrow(noaa)`) with missing values for the variable `snwd` (Snow depth (mm))

* There are `r sum(is.na(noaa$tmax))` observations (proportion: `r sum(is.na(noaa$tmax))/nrow(noaa)`) with missing values for the variable `tmax` (Maximum temperature (tenths of degrees C)) -- that's nearly half of all observations that have missing values for `tmax`!

* There are `r sum(is.na(noaa$tmin))` observations (proportion: `r sum(is.na(noaa$tmin))/nrow(noaa)`) with missing values for the variable `tmin` (Minimum temperature (tenths of degrees C)) -- that's nearly half of all observations that have missing values for `tmin`!

* There are no observations with missing values for the variables `id` and `date` (thankfully!)


### Data cleaning the NY NOAA (`noaa`) dataset

Do some data cleaning. First we create separate variables for year, month, and dayf. We also ensure that observations for temperature, precipitation, and snowfall are given in reasonable units by dividing the values for the temperature variables by 10 (changing from tenths of degree Celsius to degrees Celcius), and by dividing the values for precipitation and snowfall by 10 (changing from mm to cm).

```{r}
noaa = noaa %>%
  janitor::clean_names() %>%
  mutate(year = lubridate::year(date),
         month = lubridate::month(date, label = TRUE, abbr = FALSE),
         day = lubridate::day(date)) %>% # Creating separate variables for year, month, and day
  mutate(prcp = prcp/10,
         snow = snow/10,
         snwd = snwd/10,
         tmax = as.numeric(tmax) / 10,
         tmin = as.numeric(tmin) / 10) %>% # Ensuring reasonable units for the weather data variables
  select(id, year, month, day, everything(), -date) # Rearranging variable order so the weather data variables come last

noaa
```
### Most common observed snowfall totals

```{r Finding_most_common_snowfall}
snow_count = noaa %>% count(snow) %>% arrange(desc(n))
snow_count$snow[1]
```

**For snowfall, the most commonly observed values seem to be `r snow_count$snow[1]` followed closely by `r snow_count$snow[2]`, probably because there are many weather station locations where it doesn't snow at all (basically, anywhere where the climate is mild/warm year-round or doesn't support snowfall).**



### Two-panel plot of average max temperature in January and July, in each station across years (`noaa` dataset)

Make a two-panel plot showing the average max temperature in January and in July in each station across years. First we get the data ready to be used in a ggplot by filtering so only observations made in January and July are kept, and creating a new variable of average max temperatures (grouped by station id, year, and month). Then we use ggplot to create a two-panel plot.

```{r Plotting-average-max-temps-Jan-July}
noaa %>% 
  filter(month == c("January", "July")) %>% # Only keep data collected in January and July
  drop_na(tmax) %>% # Drop observations missing tmax data
  group_by(year, month, id) %>% # Group data by year, month, and id
  summarise(mean_tmax = mean(tmax, na.rm = TRUE)) %>% # Create new variable for average max temperatures, grouped by station id, year, and month
  
  ggplot(aes(x = year, y = mean_tmax), group = id) + # Create a scatterplot of average max temperatures in each station across the years
  geom_point(size = 0.5, alpha = 0.3) +
  facet_grid(. ~ month) + # Two-panel plot separating January and July
  labs(title = "Average max temperature in each station across the years (January vs July)",
       x = "Year",
       y = "Average max temperature (in degrees Celsius)") +
  theme_classic()
```

It seems pretty clear that although there are variations in max temperature within each year, across the years, and across different weather stations, there is generally a clear average max temperature difference between observations made in January vs July, with temperatures in January hovering around 15-20 degrees Celsius lower than average max temperatures in July across the years. Most of the January average max temperatures fall below 10 degrees Celsius; while most of the July average max temperatures fall above 20 degrees Celsius. There are some outliers at both extremes for both the January and July data across the years; for instance, in a few years both months have max average temperatures which hover between 10-20 degrees Celsius (relatively high for January and low for July). It's difficult to tell any overall trends in average max temperature across the years from these plots, if only because there is still quite a bit of variation within each year and between years.


### Plot of tmax and tmin (`noaa` dataset)

Make a plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option). 

```{r tmax_tmin_plot}
tmax_tmin = noaa %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex() +
  labs(x = "Minimum temperature (C)",
       y = "Maximum temperature (C)",
       title = "tmax vs tmin") +
  theme_classic()

tmax_tmin
```


### Plot of snowfall distribution, separately by year (`noaa` dataset)

Make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.
```{r snowfall_by_year_plot}
snowfall = noaa %>% 
  filter(snow > 0 & snow < 100) %>% # Only keep snowfall values greater than 0 and less than 100
  mutate(year = as.character(year)) %>% # Change year to character variable, so we can use it as a categorical variable in geom_density_ridges
  ggplot(aes(x = snow, y = year)) +
  geom_density_ridges() +
  labs(title = "Snowfall distribution by year",
       caption = "For snowfall values greater than 0 cm and less than 100 cm",
       x = "Snowfall (cm)",
       y = "Year") +
  theme_classic()

snowfall
```

### Two-panel plot of tmax vs tmin and snowfall distribution by year

Now we can combine the previous two plots, and create a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.


```{r two-panel-tmax-tmin-snowfall-plot}
tmax_tmin_snowfall = (tmax_tmin + snowfall)
tmax_tmin_snowfall
```

